---
title: "Project Proposal: Weather and Bicycle Counts in Munich"
author:
  - "Shangxun Liu (GitHub: lsx17)"
  - "Haoyu Meng (GitHub: MHY112928)"
  - "Zihan Wang (GitHub: Wzh927)"
date: "2026-06-08"
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
```

```{r}
#| label: load-data
raw_files <- list.files(
  "data/raw",
  pattern = "^rad_[0-9]{4}_daily\\.csv$",
  full.names = TRUE
)
processed_path <- "data/processed/bicycle_counts_clean.csv"

if (length(raw_files) == 0L) {
  source("code/01_download.R")
}

if (!file.exists(processed_path)) {
  source("code/02_clean.R")
}

bicycle_counts <- read_csv(processed_path, show_col_types = FALSE)
```

## Dataset & Research Questions

### Source and licence

This project uses the public dataset **Daten der Raddauerzaehlstellen Muenchen - Jahreszahlen** from the [City of Munich Open Data Portal](https://opendata.muenchen.de/dataset/022a11ff-4dcb-4f03-b7dd-a6c94a094587/). We found it through the Datengartln dataset page for the same dataset. The dataset contains bicycle counts from permanent counting stations in Munich.

The data is published under **Datenlizenz Deutschland Namensnennung 2.0**. This licence allows reuse, modification, and redistribution as long as attribution is provided. We will cite the City of Munich Open Data Portal and the responsible municipal data provider in all rendered outputs.

For the proposal we use the daily values with weather variables from **2021 to 2024**. We do not use 2025 because it is incomplete in the current resource list. The daily files are more suitable than the 15-minute files for the first project proposal because they already combine bicycle counts with daily weather indicators.

### What is in the dataset?

```{r}
#| label: data-size
n_rows <- nrow(bicycle_counts)
n_cols <- ncol(bicycle_counts)
date_range <- range(bicycle_counts$date, na.rm = TRUE)
station_levels <- sort(unique(bicycle_counts$station))

tibble(
  rows = n_rows,
  columns = n_cols,
  first_day = date_range[1],
  last_day = date_range[2],
  stations = paste(station_levels, collapse = ", ")
)
```

Each row represents one day at one bicycle counting station. The key variables are:

- `date`, `year`, `month`, and `weekday`: time information for each station-day.
- `station`: the bicycle counting station.
- `direction_1`, `direction_2`, and `total_count`: bicycle counts by direction and in total.
- `min_temp`, `max_temp`, and `mean_temp`: daily temperature indicators.
- `precipitation`, `cloud_cover`, and `sunshine_hours`: daily weather variables.
- `comment`: notes from the source file where available.

### Research questions

1. **How do daily bicycle counts in Munich vary across seasons and years?**  
   This descriptive question examines whether bicycle counts follow clear annual and seasonal patterns between 2021 and 2024.

2. **How are daily weather conditions related to bicycle counts?**  
   This is mainly an exploratory question. We want to see whether bicycle counts are higher on warmer or sunnier days and lower on rainy days. Because we are still R beginners and the data is observational, we will describe patterns carefully instead of making strong causal claims.

3. **Do bicycle-count patterns differ across counting stations?**  
   This comparative question checks whether some stations usually count more bicycles than others and whether the visible patterns look similar across stations.

### Context and target audience

The project is framed as a short student report about everyday cycling in Munich. Our main audience is other students and readers who are interested in city life and mobility, but who do not necessarily have a technical background. A Quarto HTML report is useful because it lets us show the text, R code, tables, and plots in one document.

## Initial Data Analysis

### Structure

```{r}
#| label: glimpse-data
glimpse(bicycle_counts)
```

### Summary statistics

```{r}
#| label: summary-statistics
bicycle_counts |>
  summarise(
    observations = n(),
    missing_total_count = sum(is.na(total_count)),
    min_daily_count = min(total_count, na.rm = TRUE),
    median_daily_count = median(total_count, na.rm = TRUE),
    mean_daily_count = round(mean(total_count, na.rm = TRUE), 1),
    max_daily_count = max(total_count, na.rm = TRUE),
    mean_temperature = round(mean(mean_temp, na.rm = TRUE), 1),
    mean_precipitation = round(mean(precipitation, na.rm = TRUE), 2)
  )
```

```{r}
#| label: station-summary
bicycle_counts |>
  group_by(station) |>
  summarise(
    observations = n(),
    first_day = min(date, na.rm = TRUE),
    last_day = max(date, na.rm = TRUE),
    missing_total_count = sum(is.na(total_count)),
    median_daily_count = median(total_count, na.rm = TRUE),
    mean_daily_count = round(mean(total_count, na.rm = TRUE), 1),
    .groups = "drop"
  ) |>
  arrange(desc(mean_daily_count))
```

### Plot 1: distribution of daily bicycle counts

```{r}
#| label: plot-distribution
#| fig-cap: "Distribution of daily bicycle counts across station-days."
bicycle_counts |>
  filter(!is.na(total_count)) |>
  ggplot(aes(x = total_count)) +
  geom_histogram(bins = 40, fill = "#2A7F62", colour = "white") +
  labs(
    title = "Daily bicycle counts are strongly right-skewed",
    x = "Daily bicycle count",
    y = "Number of station-days"
  ) +
  theme_minimal()
```

### Plot 2: temperature and bicycle counts

```{r}
#| label: plot-weather-temperature
#| fig-cap: "Relationship between mean daily temperature and bicycle counts."
bicycle_counts |>
  filter(!is.na(total_count), !is.na(mean_temp)) |>
  ggplot(aes(x = mean_temp, y = total_count, colour = station)) +
  geom_point(alpha = 0.35, size = 1.2) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, linewidth = 0.8) +
  labs(
    title = "Warmer days tend to have higher bicycle counts",
    x = "Mean daily temperature (C)",
    y = "Daily bicycle count",
    colour = "Station"
  ) +
  theme_minimal()
```

### Data quality notes

The first screening suggests several data quality and interpretation issues:

- The source dataset is organised by year, so the download script must combine several CSV files and keep a record of the source file.
- Date formats are not fully consistent across years. Some files use date-like strings with hyphens, while newer files use dots. The cleaning script therefore reads dates as text and parses them explicitly.
- Station availability changes over time. The source notes that some stations were affected by construction, damage, or technical defects. For example, some 2024 station records are missing or not fully reliable.
- The data contains station-level counts, not all bicycle trips in Munich. We therefore cannot say how many bicycle trips happened in the whole city.
- Weather variables are useful for first exploratory plots, but they are not enough to prove cause and effect. Other factors such as holidays, school vacations, public events, construction, or changes in infrastructure may also matter.

## Analysis Plan

### Research question 1: seasonality and time trends

We will use `date`, `year`, `month`, `weekday`, `station`, and `total_count`. Our main approach will be exploratory data analysis with line plots, monthly summaries, weekday comparisons, and simple station-level plots. The key cleaning steps are downloading annual daily CSVs, standardising date formats, and combining all years into one station-day dataset.

### Research question 2: weather and bicycle counts

We will use `mean_temp`, `precipitation`, `cloud_cover`, `sunshine_hours`, and `total_count`. First, we will use scatterplots and grouped summaries to describe visible relationships. If we feel confident enough later in the project, we may also try a simple linear regression, but the main goal is to understand and explain the patterns rather than to build a complicated model. Results will be interpreted as associations, not causal effects.

### Research question 3: station differences

We will compare `total_count` across `station` using grouped summary statistics and station-specific plots. If the final report has enough time, we may compare whether the weather plots look different by station, but this will be exploratory.

## Workflow Organisation

### Code style

We will try to follow the tidyverse style guide: clear object names, pipe-based transformations, one meaningful step per line, and lowercase snake_case names for new variables. Because we are still learning R, we will check each other's scripts before submitting and may use `styler` to make the formatting more consistent. Date parsing and source-file handling are kept explicit because they are important reproducibility steps for this dataset.

### Packages

The initial project relies on `readr`, `dplyr`, `ggplot2`, and `lubridate`. These packages are enough for downloading, cleaning, summarising, and plotting the data. We may add one or two packages later if we really need them, but we want to keep the project understandable for all group members. We do not use `renv` yet, but after the W9 reproducibility session we will decide whether to add it for the final project.

### Git workflow

The repository will stay on GitHub. The main branch should contain working material. For now, we will keep the workflow simple: one person can work on data cleaning, one on plots, and one on writing. If we make larger changes, we can use separate branches such as `data-cleaning`, `weather-eda`, and `writeup-analysis-plan`, then merge them after review. Commit messages should describe what changed, not only which file changed.

### `.gitignore`

The project excludes local R session files, Quarto cache files, rendered support files, downloaded raw CSV files, processed CSV files, and possible secrets such as `.env` files. The repository includes scripts that download and recreate the data, so we do not need to store large generated files on GitHub. If the final project requires a fixed data snapshot, we will document that decision clearly.
